Computing Entropies With Nested Sampling

### 2017-05-21 paper/ms.tex
Bit of tidying

% Abstract (Do not use inserted blank lines, i.e. \\)
\abstract{The Shannon entropy, and related quantities such as mutual
information, can be used to quantify uncertainty and relevance.
However, in practice, it can be difficult to compute these quantities
for arbitrary probability distributions, particularly if some probability
mass functions or densities cannot be evaluated. This paper introduces a
computational approach to evaluating entropies of probability distributions
that can be sampled. The method is based on Nested Sampling, and is related to Approximate Bayesian Computation. I demonstrate the method
on a simple gaussian example where the key quantities are known analytically,
and an experimental design example about scheduling observations in order
to measure the period of an oscillating signal.}

@@

Throughout this paper, I use the compact
`overloaded' notation for probability distributions favoured by
Bayesians, writing $p(x)$ (for either a probability mass function
or a probability density function) instead
of the more correct but cumbersome $P(X=x)$ or $f_X(x)$.
Probability distributions are implicitly conditional on some prior
information, which is omitted from the notation unless necessary.
All logarithms are written $\log$ and any base can be used, unless otherwise
specified (for example by writing $\ln$ or $\log_{10}$). Any numerical
values given for the value of specific entropies are in {\em nats}, i.e.,
the natural logarithm was used.

I prefer to write the entropy of $x$ as $H_x$ rather than $H(x)$
(or $H(X)$ as is often seen), because the latter can wrongly imply that
the entropy is a property of the value of $x$ itself, rather than
our probability distribution for it.

@@

In practical applications of Bayesian inference, the following operations
are usually feasible, and are assumed to be possible at low computational
cost:
\begin{enumerate}
\item Points can be generated from the prior $p(\x)$;
\item Simulated datasets can be generated from $p(\d | \x)$ for a given value of $\x$;
\item The likelihood, $p(\d | \x)$ can be evaluated cheaply for any $\d$ and $\x$.
\end{enumerate}

@@

This is difficult to compute because the expression
for the posterior distribution
\begin{align}
p(\x | \d) &= \frac{p(\x)p(\d | \x)}{p(\d)}
\end{align}
contains the marginal likelihood or `evidence'
\begin{align}
p(\d) &= \int p(\x) p(\d | \x) \, d\x.
\end{align}

It is important to distinguish between $H_{\x | \d}$ and the
entropy of $\x$ given a particular value of $\d$, which might be written
$H_{\x | \d=\d_0}$.
The former measures the degree to which one question answers
another, so is a function of two questions. The latter measures the
degree to which a {\em statement} answered a question, and is a function
of a question and a statement.

\subsection{Mutual information}
The mutual information is another way of describing the relevance of the
data to the parameters. Its definition, and relation to other quantities, is
\begin{align}
I_{\x, \d} &= \sum_{\x} \sum_{\d} p(\x, \d)
                       \log\left[\frac{p(\x, \d)}{p(\x)p(\d)}\right]\\
           &= H_{\x} + H_{\d} - H_{\x, \d}\\
           &= H_{\d} - H_{\d | \x}\\
           &= H_{\x} - H_{\x | \d}.
\end{align}
It can also be written in terms of the prior, likelihood, and evidence,
as
\begin{align}
I_{\x, \d} &= \sum_{\x} \sum_{\d} p(\x)p(\d | \x)
              \left[\log p(\d | \x) - \log p(\d)\right],
\end{align}
i.e., the mutual information is the prior expected value of the
log likelihood minus the log evidence.
As with the conditional entropy, we see that the difficulty appears in the
form of the log of the marginal likelihood,
$\log p(\d) = \log \sum_{\x} p(\x)p(\d | \x)$.

@@

Nested Sampling \citep[NS][]{skilling2006nested} is an algorithm whose aim is
to calculate the marginal likelihood
\begin{align}
Z &= \int \pi(\x) L(\x) \, d\x
\end{align}
where $\x$ is the unknown parameter(s), $\pi$ is the prior distribution,
and $L$ is the likelihood function. It does this evolving a population of
$N$ particles in the parameter space.
Defining $X(\ell)$ as the amount of prior mass with likelihood greater than
some threshold $\ell$,
\begin{align}
X(\ell) &= \int \pi(\x) \mathds{1}\left(L(\x) > \ell\right) \, d\x,
\end{align}

@@

When $N=1$, the sequence of $\ln X$ values is a Poisson process with rate
1. \citet{Walter2015} showed how this view of the NS sequence of points
can be used to construct a
version of NS that produces unbiased estimates of the marginal likelihood.
However, it can also be used to construct an unbiased estimator of
a log-probability, which is more relevant to information theoretic
quantities.

>>>>

% Abstract (Do not use inserted blank lines, i.e. \\)
\abstract{The Shannon entropy, and related quantities such as mutual
information, can be used to quantify uncertainty and relevance.
However, in practice, it can be difficult to compute these quantities
for arbitrary probability distributions, particularly if the probability
mass functions or densities cannot be evaluated. This paper introduces a
computational approach to evaluating entropies of probability distributions
that can only be sampled. The method is based on Nested Sampling,
and is related to Approximate Bayesian Computation. I demonstrate the method
on a simple gaussian example where the key quantities are known analytically,
and an experimental design example about scheduling observations in order
to measure the period of an oscillating signal.}

@@

Throughout this paper, I use the compact
`overloaded' notation for probability distributions favoured by many
Bayesian writers, writing $p(x)$ (for either a probability mass function
or a probability density function) instead
of the more correct but cumbersome $P(X=x)$ or $f_X(x)$.
Probability distributions are implicitly conditional on some prior
information, which is omitted from the notation unless necessary.
All logarithms are written $\log$ and any base can be used, unless otherwise
specified (for example by writing $\ln$ or $\log_{10}$). Any numerical
values given for the value of specific entropies are in {\em nats}, i.e.,
the natural logarithm was used.

I prefer to write the entropy of $x$ as $H_x$ rather than $H(x)$
(or $H(X)$ as is often seen), because the latter can wrongly imply that
the entropy is a property of the value of $x$ itself, rather than
the probability distribution we have assigned to it.

@@

In practical applications, the following operations
are usually feasible and have a low computational cost:
\begin{enumerate}
  \item Points can be generated from the prior $p(\x)$;
  \item Simulated datasets can be generated from $p(\d | \x)$ for any
        given value of $\x$;
  \item The likelihood, $p(\d | \x)$, can be evaluated cheaply for any
        $\d$ and $\x$.
\end{enumerate}
Throughout this paper I assume these operations are available and inexpensive.

@@

Suppose we wanted to compute $H_{\x | \d}$.
This would be difficult because the expression
for the posterior distribution
\begin{align}
p(\x | \d) &= \frac{p(\x)p(\d | \x)}{p(\d)}
\end{align}
contains the marginal likelihood integral:
\begin{align}
p(\d) &= \int p(\x) p(\d | \x) \, d\x
\end{align}
also known as the `evidence'. In this paper I use the term evidence because
for brevity.

It is important to distinguish between $H_{\x | \d}$ and the
entropy of $\x$ given a particular value of $\d$, which might be written
$H_{\x | \d=\d_0}$.
The former measures the degree to which one question answers
another (ex ante), so is a function of two questions. The latter measures the
degree to which a {\em statement} answered a question (ex post),
and is a function of a question and a statement.

\subsection{Mutual information}
The mutual information is another way of describing the relevance of the
data to the parameters. Its definition, and relation to other quantities, is
\begin{align}
I_{\x, \d} &= \sum_{\x} \sum_{\d} p(\x, \d)
                       \log\left[\frac{p(\x, \d)}{p(\x)p(\d)}\right]\\
           &= H_{\x} + H_{\d} - H_{\x, \d}\\
           &= H_{\d} - H_{\d | \x}\\
           &= H_{\x} - H_{\x | \d}.
\end{align}
It can also be written in terms of the prior, likelihood, and evidence,
as
\begin{align}
I_{\x, \d} &= \sum_{\x} \sum_{\d} p(\x)p(\d | \x)
              \left[\log p(\d | \x) - \log p(\d)\right],
\end{align}
i.e., the mutual information is the prior expected value of the
log likelihood minus the log evidence.
As with the conditional entropy, we see that the difficulty appears in the
form of the log evidence,
$\log p(\d) = \log \sum_{\x} p(\x)p(\d | \x)$.

@@

Nested Sampling \citep[NS][]{skilling2006nested} is an algorithm whose aim is
to calculate the evidence
\begin{align}
Z &= \int \pi(\x) L(\x) \, d\x
\end{align}
where $\x$ is the unknown parameter(s), $\pi$ is the prior distribution,
and $L$ is the likelihood function. It does this evolving a population of
$N$ particles in the parameter space.
Defining $X(\ell)$ as the amount of prior mass with likelihood greater than
some threshold $\ell$,
\begin{align}
X(\ell) &= \int \pi(\x) \mathds{1}\left(L(\x) > \ell\right) \, d\x,
\end{align}

@@

When $N=1$, the sequence of $\ln X$ values is a Poisson process with rate
1. \citet{Walter2015} showed how this view of the NS sequence of points
can be used to construct a
version of NS that produces unbiased estimates of the evidence.
However, it can also be used to construct an unbiased estimator of
a log-probability, which is more relevant to information theoretic
quantities.
